Revisiting the Relationship between Short-Chain Fatty Acids, the Microbiota, and Colorectal Tumors

### 2019-03-19 submission/manuscript.Rmd
Add code to classification paragraph

```{r, classification}
```

**Combining SCFA and microbiome data does not improve the ability to diagnose individual as having adenomas or carcinomas.** We previously found that binning 16S rRNA gene sequence data into operational taxonomic units based on 97% similarity or into genera enabled us to classify individuals as having adenomas or carcinomas using Random Forest machine learning models. We repeated that analysis but added the concentration of the individual or pooled SCFAs as possible features to train a model. A model trained on the individual or pool SCFAs alone had a median area under the receiver operator characteristic curve (AUROC) of **0.XX** (IQR=**0.XX**). When we trained the models with the SCFAs concentrations and OTU or genus-level relative abundances (AUROC~OTU+SCFA~=**0.XX**, AUROC~genus+SCFA~=**0.XX**), the AUROC values were no different from the models trained without the SCFA concentrations (AUROC~OTU~=**0.XX**, AUROC~genus~=**0.XX**; Figure 2A). We also trained models using a smaller dataset that generated shotgun metagenomic sequencing data from a subset of our cohort (N~normal~=**XX**, N~adenoma~=**XX**, and N~cancer~=**XX**). We binned genes extracted from the assembled metagenomes into operational protein families (OPFs) or KEGG categories [**XXXXXX**]. Again, the performance of the models trained with the meteagenomic data did not improve when the SCFA concentrations were added as possible features when training the model (Figure 2A). These data demonstrate that knowledge of the SCFA profile from a patient's fecal sample does not improve the ability to diagnose a colonic lesion.

>>>>

```{r, classification}
random <- read_tsv("../data/rf/classification_SCFA_to_random.tsv")
stopifnot(random$p.value < 0.001)

w_wo_compare <- read_tsv("../data/rf/classification_w_wo_SCFA.tsv")

min_p_compare_tax <- w_wo_compare %>%
	filter(class=="adenoma" | class == "cancer") %>%
	filter(microbiome =="otu" | microbiome == "genus") %>%
  select(p.value) %>%
	min()

min_p_compare_metag <- w_wo_compare %>%
	filter(class=="adenoma" | class == "cancer") %>%
	filter(microbiome =="opf" | microbiome == "kegg") %>%
	select(p.value) %>%
	min()

n_s <- read_tsv("../data/metagenome/metag.sample.counts")
```

**Combining SCFA and microbiome data does not improve the ability to diagnose individual as having adenomas or carcinomas.** We previously found that binning 16S rRNA gene sequence data into operational taxonomic units based on 97% similarity or into genera enabled us to classify individuals as having adenomas or carcinomas using Random Forest machine learning models [@Baxter2016; @Sze2018]. We repeated that analysis but added the concentration of the individual or pooled SCFAs as possible features to train a model. Models trained using SCFAs to classify individuals as having adenomas or carcinomas rather than normal colons had median areas under the receiver operator characteristic curve (AUROC) that were significantly greater than 0.5 (P~adenoma~<0.001 and P~carcinoma~<0.001); however, the AUROC values to detect the presence of adenomas or carcinomas were only `r random %>% filter(class=="adenoma") %>% pull(median)` and `r random %>% filter(class=="cancer") %>% pull(median)`, respectively (Figure 2A). When we trained the models with the SCFAs concentrations and OTU or genus-level relative abundances the AUROC values were not significantly different from the models trained without the SCFA concentrations (P>`r min_p_compare_tax`; Figure 2A). We also trained models using a smaller dataset that generated shotgun metagenomic sequencing data from a subset of our cohort (N~normal~=`r n_s %>% filter(dx == "normal") %>% pull(n)`, N~adenoma~=`r n_s %>% filter(dx == "adenoma") %>% pull(n)`, and N~cancer~=`r n_s %>% filter(dx == "cancer") %>% pull(n)`) [@Hannigan2017]. We binned genes extracted from the assembled metagenomes into operational protein families (OPFs) or KEGG categories. Again, the performance of the models trained with the meteagenomic data did not improve when the SCFA concentrations were added as possible features when training the model (P>`r min_p_compare_metag`; Figure 2A). These data demonstrate that knowledge of the SCFA profile from a patient's fecal sample does not improve the ability to diagnose a colonic lesion.

### 2019-03-19 submission/manuscript.tex
Add code to classification paragraph

\textbf{Combining SCFA and microbiome data does not improve the ability
to diagnose individual as having adenomas or carcinomas.} We previously
found that binning 16S rRNA gene sequence data into operational
taxonomic units based on 97\% similarity or into genera enabled us to
classify individuals as having adenomas or carcinomas using Random
Forest machine learning models. We repeated that analysis but added the
concentration of the individual or pooled SCFAs as possible features to
train a model. A model trained on the individual or pool SCFAs alone had
a median area under the receiver operator characteristic curve (AUROC)
of \textbf{0.XX} (IQR=\textbf{0.XX}). When we trained the models with
the SCFAs concentrations and OTU or genus-level relative abundances
(AUROC\textsubscript{OTU+SCFA}=\textbf{0.XX},
AUROC\textsubscript{genus+SCFA}=\textbf{0.XX}), the AUROC values were no
different from the models trained without the SCFA concentrations
(AUROC\textsubscript{OTU}=\textbf{0.XX},
AUROC\textsubscript{genus}=\textbf{0.XX}; Figure 2A). We also trained
models using a smaller dataset that generated shotgun metagenomic
sequencing data from a subset of our cohort
(N\textsubscript{normal}=\textbf{XX},
N\textsubscript{adenoma}=\textbf{XX}, and
N\textsubscript{cancer}=\textbf{XX}). We binned genes extracted from the
assembled metagenomes into operational protein families (OPFs) or KEGG
categories {[}\textbf{XXXXXX}{]}. Again, the performance of the models
trained with the meteagenomic data did not improve when the SCFA
concentrations were added as possible features when training the model
(Figure 2A). These data demonstrate that knowledge of the SCFA profile
from a patient's fecal sample does not improve the ability to diagnose a
colonic lesion.

@@

\textbf{Conclusion.}

\begin{itemize}
\tightlist
\item
  Unlikely that SCFAs are the primary mechanism that limits
  tumorigenesis
\item
  Limitations

  \begin{itemize}
  \tightlist
  \item
    Observing the community once the tumor is developed
  \item
    Not looking at mucosal communities
  \end{itemize}
\item
  Need to identify other possible mechanisms that drive and prevent
  tumorigenesis
\item
  Supports evidence that fiber consumption has a null effect on
  tumorigenesis
\end{itemize}

>>>>

\textbf{Combining SCFA and microbiome data does not improve the ability
to diagnose individual as having adenomas or carcinomas.} We previously
found that binning 16S rRNA gene sequence data into operational
taxonomic units based on 97\% similarity or into genera enabled us to
classify individuals as having adenomas or carcinomas using Random
Forest machine learning models (13, 14). We repeated that analysis but
added the concentration of the individual or pooled SCFAs as possible
features to train a model. Models trained using SCFAs to classify
individuals as having adenomas or carcinomas rather than normal colons
had median areas under the receiver operator characteristic curve
(AUROC) that were significantly greater than 0.5
(P\textsubscript{adenoma}\textless{}0.001 and
P\textsubscript{carcinoma}\textless{}0.001); however, the AUROC values
to detect the presence of adenomas or carcinomas were only 0.54 and
0.55, respectively (Figure 2A). When we trained the models with the
SCFAs concentrations and OTU or genus-level relative abundances the
AUROC values were not significantly different from the models trained
without the SCFA concentrations (P\textgreater{}0.21; Figure 2A). We
also trained models using a smaller dataset that generated shotgun
metagenomic sequencing data from a subset of our cohort
(N\textsubscript{normal}=27, N\textsubscript{adenoma}=25, and
N\textsubscript{cancer}=26) (16). We binned genes extracted from the
assembled metagenomes into operational protein families (OPFs) or KEGG
categories. Again, the performance of the models trained with the
meteagenomic data did not improve when the SCFA concentrations were
added as possible features when training the model (P\textgreater{}0.24;
Figure 2A). These data demonstrate that knowledge of the SCFA profile
from a patient's fecal sample does not improve the ability to diagnose a
colonic lesion.

@@

\textbf{Conclusion.} Our data indicate that fecal SCFA concentrations
are not associated with the presence of adenomas or carcinomas and that
they provide weak predictive power to improve the ability to diagnose
someone with one of these lesions. Furthermore, knowledge of the
taxonomic and genetic structure of gut microbiota was not predictive of
individual or pooled SCFA concentrations. These results complement
existing literature that suggest that fiber consumption and the
production of SCFAs are unable to prevent the risk of developing colonic
tumors. It is important to note that our analysis concerned fecal SCFA
concentrations and microbiome characterization and that observations
along the mucosa near the site of lesions may provide a stronger
association. Regardless, given the growing literature in this area, it
is unlikely that SCFAs are the primary mechanism that limits
tumorigenesis. This may be a cautionary result to temper enthusiasm for
SCFAs as a biomarker of gut health more generally. Going forward it is
critical to develop additional hypotheses for how the microbiome and
host interact to drive tumorigenesis to better understand the disease
process and identify biomarkers that will allow early detection of
tumors.
